# Learning R and Bioconductor {#learning-r-and-more}

In this section, we outline various resources for learning R and Bioconductor. We provide a brief set of instructions for installing R on your own machine, and then cover how to get help for functions, packages, and Bioconductor-specific resources for learning more. 

## The Benefits of R and Bioconductor

[R](http://www.r-project.org/) is a high-level programming language that was initially designed for statistical applications. While there is much to be said [about R](https://www.r-project.org/about.html) as a programming language, one of the key advantages of using R is that it is highly extensible through _packages_. Packages are collections of functions, data, and documentation that extends the capabilities of base R. The ease of development and distribution of packages for R has made it a rich environment for many fields of study and application.

One of the primary ways in which packages are distributed is through centralized repositories. The first R repository a user typically runs into is the [Comprehensive R Archive Network](https://cran.r-project.org/mirrors.html) (CRAN), which hosts over 13,000 packages to date, and is home to many of the most popular R packages. 

Similar to CRAN, [Bioconductor](https://bioconductor.org) is a repository of R packages as well. However, whereas CRAN is a general purpose repository, Bioconductor focuses on software tailored for genomic analysis. Furthermore, Bioconductor has stricter requirements for a package to be accepted into the repository. Of particular interest to us is the inclusion of [high quality documentation](#learning-more) and the use of common [data infrastructure](#data-infrastructure) to promote package interoperability.

In order to use these packages from CRAN and Bioconductor, and start programming with R to follow along in these workflows, some knowledge of R is helpful. Here we outline resources to guide you through learning the basics. 


## Learning R Online {#getting-started-with-r}

To learn more about programming with R, we highly recommend checking out the online courses offered by [Datacamp](https://www.datacamp.com/), which includes both introductory and advanced courses within the [R track](https://www.datacamp.com/courses/tech:r). Datacamp is all online with many free courses, with videos and a code editor/console that promotes an interactive learning experience. What we like about Datacamp is that it is more focused on topics and programming paradigms that center around data science, which is especially helpful for getting started with R. 

Beyond just Datacamp, a mainstay resource for learning R is the [R for Data Science](https://r4ds.had.co.nz/) book. This book illustrates R programming through the exploration of various data science concepts - transformation, visualization, exploration, and more. 


## Running R Locally

While learning R through online resources is a great way to start with R, as it requires minimal knowledge to start up, at some point, it will be desirable to have a local installation - on your own hardware - of R. This will allow you to install and maintain your own software and code, and furthermore allow you to create a personalized workspace.

### Installing R

Prior to getting started with this book, some prior programming experience with R is helpful. Check out the [_Learning R and More_](#learning-r-and-more) chapter for a list of resources to get started with R and other useful tools for bioinformatic analysis.

To follow along with the analysis workflows in this book on your personal computer, it is first necessary to install the [R](http://www.r-project.org/) programming language. Additionally, we recommend a graphical user interface such as [RStudio](http://www.rstudio.com/download) for programming in R and visualization. RStudio features many helpful tools, such as code completion and an interactive data viewer to name but two. For more details, please see the online book [_R for Data Science_ prerequisites](https://r4ds.had.co.nz/introduction.html#prerequisites) section for more information about installing R and using RStudio.

#### For MacOS/Linux Users

A special note for MacOS/Linux users: we highly recommend using a package manager to manage your R installation. This differs across different Linux distributions, but for MacOS we highly recommend the [Homebrew](https://brew.sh/) package manager. Follow the website directions to install homebrew, and install R via the commandline with `brew install R`, and it will automatically configure your installation for you. Upgrading to new R versions can be done by running `brew upgrade`.

### Installing R & Bioconductor Packages

After installing R, the next step is to install R packages. In the R console, you can install packages from CRAN via the `install.packages()` function. In order to [install Bioconductor packages](https://www.bioconductor.org/install/), we will first need the _BiocManager_ package which is hosted on CRAN. This can be done by running:

```{r, eval=FALSE}
install.packages("BiocManager")
```

The _BiocManager_ package makes it easy to install packages from the Bioconductor repository. For example, to install the [_SingleCellExperiment_](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) package, we run:

```{r, eval=FALSE}
## the command below is a one-line shortcut for:
## library(BiocManager)
## install("SingleCellExperiment")
BiocManager::install("SingleCellExperiment")
```

Throughout the book, we can load packages via the `library()` function, which by convention usually comes at the top of scripts to alert readers as to what packages are required. For example, to load the _SingleCellExperiment_ package, we run:

```{r, eval=FALSE}
library(SingleCellExperiment)
```

Many packages will be referenced throughout the book within the workflows, and similar to the above, can be installed using the `BiocManager::install()` function. 


## Getting Help In (and Out) of R

One of the most helpful parts of R is being able to get help _inside_ of R. For example, to get the manual associated with a function, class, dataset, or package, you can prepend the code of interest with a `?` to retrieve the relevant help page. For example, to get information about the `data.frame()` function, the _SingleCellExperiment_ class, the in-built _iris_ dataset, or for the _BiocManager_ package, you can type:

```{r, eval=FALSE}
?data.frame
?SingleCellExperiment
?iris
?BiocManager
```

Beyond the R console, there are myriad online resources to get help. The R for Data Science book has a great section dedicated to looking for help [outside of R](https://r4ds.had.co.nz/introduction.html#getting-help-and-learning-more). In particular, [Stackoverflow's R tag](https://stackoverflow.com/questions/tagged/r) is a helpful resource for asking and exploring general R programming questions.


## Bioconductor Help {#bioconductor-documentation}

One of the key tenets of Bioconductor software that makes it stand out from CRAN is the required documentation of packages and workflows. In addition, Bioconductor hosts a Bioconductor-specific support site that has grown into a valuable resource of its own, thanks to the work of dedicated volunteers. Here we briefly dive into the


Bioconductor package vignettes and reference manuals
Bioconductor BiocViews and workflows
Bioconductor forum
Stackoverflow and Google for more generic R programming advice
