# Using R & Bioconductor {#using-r-and-bioconductor}

## A Brief Overview

[R](http://www.r-project.org/) is a high-level programming language that was initially designed for statistical applications. While there is much to be said [about R](https://www.r-project.org/about.html) as a programming language, one of the key advantages of using R is that it is highly extensible through _packages_. Packages are collections of functions, data, and documentation that extends the capabilities of base R. The ease of development and distribution of packages for R has made it a rich environment for many fields of study and application.

One of the primary ways in which packages are distributed is through centralized repositories. The first R repository a user typically runs into is the [Comprehensive R Archive Network](https://cran.r-project.org/mirrors.html) (CRAN), which hosts over 13,000 packages to date, and is home to many of the most popular R packages. 

Similar to CRAN, [Bioconductor](https://bioconductor.org) is a repository of R packages as well. However, whereas CRAN is a general purpose repository, Bioconductor focuses on software tailored for genomic analysis. Furthermore, Bioconductor has stricter requirements for a package to be accepted into the repository. Of particular interest to us is the inclusion of [high quality documentation](#learning-more) and the use of common [data infrastructure](#data-infrastructure) to promote package interoperability.


## Prerequisites {#prerequisites}

### Installing R

Prior to getting started with this book, some prior programming experience with R is helpful. Check out the [_Learning R and More_](#learning-r-and-more) chapter for a list of resources to get started with R and other useful tools for bioinformatic analysis.

To follow along with the analysis workflows in this book on your personal computer, it is first necessary to install the [R](http://www.r-project.org/) programming language. Additionally, we recommend a graphical user interface such as [RStudio](http://www.rstudio.com/download) for programming in R and visualization. For more details, please see the online book [_R for Data Science_ prerequisites](https://r4ds.had.co.nz/introduction.html#prerequisites) section.

A special note for Mac OS users: we highly recommend using the [Homebrew](https://brew.sh/) package manager for managing your R installation. Follow the website directions to install homebrew, and install R via the commandline with `brew install R`, and it will automatically configure your installation for you. Upgrading to new R versions can also be done by running `brew upgrade`.

### Installing R & Bioconductor Packages

After installing R, the next step is to install R packages. In the R console, you can install packages from CRAN via the `install.packages()` function. In order to [install Bioconductor packages](https://www.bioconductor.org/install/), we will first need the _BiocManager_ package which is hosted on CRAN. This can be done by running:

```{r, eval=FALSE}
install.packages("BiocManager")
```

The _BiocManager_ package makes it easy to install packages from the Bioconductor repository. For example, to install the [_SingleCellExperiment_](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) package, we run:

```{r, eval=FALSE}
## the command below is a one-line shortcut for:
## library(BiocManager)
## install("SingleCellExperiment")
BiocManager::install("SingleCellExperiment")
```

Throughout the book, we can load packages via the `library()` function, which by convention usually comes at the top of scripts to alert readers as to what packages are required. For example, to load the _SingleCellExperiment_ package, we run:

```{r, eval=FALSE}
library(SingleCellExperiment)
```

Many packages will be referenced throughout the book within the workflows, and similar to the above, can be installed using the `BiocManager::install()` function. 


## Getting Help In (and Out) of R

One of the most helpful parts of R is being able to get help _inside_ of R. For example, to get the manual associated with a function, class, dataset, or package, you can prepend the code of interest with a `?` to retrieve the relevant help page. For example, to get information about the `data.frame()` function, the _SingleCellExperiment_ class, the in-built _iris_ dataset, or for the _BiocManager_ package, you can type:

```{r, eval=FALSE}
?data.frame
?SingleCellExperiment
?iris
?BiocManager
```

Beyond R, there are myriad online resources to get help. The R for Data Science book has a great section dedicated to looking for help [outside of R](https://r4ds.had.co.nz/introduction.html#getting-help-and-learning-more).


## Bioconductor Documentation {#bioconductor-documentation}

One of the key tenets of Bioconductor software is the rich documentation of packages and workflows. Here, we cover some of the key resources 

Bioconductor package vignettes and reference manuals
Bioconductor BiocViews and workflows
Bioconductor forum
Stackoverflow and Google for more generic R programming advice



## Becoming a Bioconductor Expert {#becoming-a-bioconductor-expert}

While [learning to use Bioconductor](#bioconductor-documentation) tools is a very welcoming experience, there is unfortunately, no central resource for navigating the plethora of gotchas and paradigms associated with developing for Bioconductor. Based on conversations with folks involved in developing for Bioconductor, much of this knowledge is hard won and fairly spread out. This however is beginning to change with more recent efforts led by the Bioconductor team, and this book represents an earnest effort towards addressing this concern.

The [BiocWorkshops](https://github.com/Bioconductor/BiocWorkshops) repository under the [Bioconductor Github](https://github.com/Bioconductor/) provides a book composed of workshops that have been hosted by Bioconductor team members and contributors. These workshops center around learning, using, and developing for Bioconductor.

A host of topics are also available via the [Learn](http://bioconductor.org/help/course-materials/) module on the Bioconductor website as well.



## Conventions used in this book

Talk about code font, code blocks, function references, objects, classes.
