# Integrating Datasets

_Author_: Stephanie Hicks

## Introduction

The purpose of this case study is to demonstrate how to integrate 
multiple scRNA-seq datasets using the R/Bioconductor package. 

## Learning objectives

* In bullet form, illustrate what the main outputs are of this vignette
* A couple of bullet points are fine

## Package requirements

Packages that are required to go through the vignette and how to install them, should largely be through BiocManager interface.

```{r preamble, eval=FALSE}
BiocManager::install(c('scater', 'scran', 'TENxPBMCData'))
```

```{r}
library(scater)
library(scran)
setwd("/users/shicks1/projects/orchestratingSingleCellAnalysis")
```

## Data

### Description of Data

Here the data should be loaded if in package form and briefly explored to illustrate the contents of the data. 

### Loading the Data

If data comes from "raw" form, then a simple pipeline for creating the processed data should be illustrated as succinctly as possible.

```{r}
library(TENxPBMCData)
tenx_pbmc3k <- TENxPBMCData('pbmc3k')
tenx_pbmc4k <- TENxPBMCData('pbmc4k')
```

Find intersection of gene names and keep only intersection 
to have same rows or features. 
```{r}
keep_genes <- intersect(rownames(tenx_pbmc3k), rownames(tenx_pbmc4k))
tenx_pbmc3k <- tenx_pbmc3k[match(keep_genes, rownames(tenx_pbmc3k)), ]
tenx_pbmc4k <- tenx_pbmc4k[na.omit(match(keep_genes, rownames(tenx_pbmc4k))), ]
```

We will also create a combined SCE object
```{r}
counts_pbmc <- cbind(counts(tenx_pbmc3k), counts(tenx_pbmc4k))
sce <- SingleCellExperiment(assays=list(counts=counts_pbmc), 
                     rowData = rowData(tenx_pbmc3k),
                     colData = rbind(colData(tenx_pbmc3k), 
                                     colData(tenx_pbmc4k)))
```


## Preprocessing

All steps required to preprocess the data into the clean expression matrix, split it up into subsections as necessary. Subsection splits should be informed by the paper (sub)sections. This is all steps prior to the main strategy of interest.


### Cell and gene quality control

Removal of "bad cells", low abundance genes. 

```{r}
sce <- calculateQCMetrics(sce)
low_lib_sce <- isOutlier(sce$log10_total_counts, type="lower", nmad=3)
low_genes_sce <- isOutlier(sce$log10_total_features_by_counts, type="lower", nmad=3)
data.frame(LowLib=sum(low_lib_sce), LowNgenes=sum(low_genes_sce))
```

```{r}
tenx_pbmc3k <- calculateQCMetrics(tenx_pbmc3k)
low_lib_tenx_pbmc3k <- isOutlier(tenx_pbmc3k$log10_total_counts, type="lower", nmad=3)
low_genes_tenx_pbmc3k <- isOutlier(tenx_pbmc3k$log10_total_features_by_counts, type="lower", nmad=3)
data.frame(LowLib=sum(low_lib_tenx_pbmc3k), 
           LowNgenes=sum(low_genes_tenx_pbmc3k))
```

```{r}
tenx_pbmc4k <- calculateQCMetrics(tenx_pbmc4k)
low_lib_tenx_pbmc4k <- isOutlier(tenx_pbmc4k$log10_total_counts, type="lower", nmad=3)
low_genes_tenx_pbmc4k <- isOutlier(tenx_pbmc4k$log10_total_features_by_counts, type="lower", nmad=3)
data.frame(LowLib=sum(low_lib_tenx_pbmc4k), 
           LowNgenes=sum(low_genes_tenx_pbmc4k))
```


```{r}
sce <- sce[,!(low_lib_sce | low_genes_sce)]
tenx_pbmc3k <- tenx_pbmc3k[,!(low_lib_tenx_pbmc3k | low_genes_tenx_pbmc3k)]
tenx_pbmc4k <- tenx_pbmc4k[,!(low_lib_tenx_pbmc4k | low_genes_tenx_pbmc4k)]
```


### Normalization

Application of normalization approach, even if its just a 
log normalization on the count data. 
Cell cycle normalization may be applicable here.

```{r}
sce <- computeSumFactors(sce)
summary(sizeFactors(sce))
```

```{r}
tenx_pbmc3k <- computeSumFactors(tenx_pbmc3k)
summary(sizeFactors(tenx_pbmc3k))
```

```{r}
tenx_pbmc4k <- computeSumFactors(tenx_pbmc4k)
summary(sizeFactors(tenx_pbmc4k))
```

```{r}
sce <- normalize(sce)
tenx_pbmc3k <- normalize(tenx_pbmc3k)
tenx_pbmc4k <- normalize(tenx_pbmc4k)
```

Save files on clusters
```{r}
saveRDS(sce, file="data/05.integrating-datasets-sce.rds")
saveRDS(tenx_pbmc3k, file="data/05.integrating-datasets-tenx_pbmc3k.rds")
saveRDS(tenx_pbmc4k, file="data/05.integrating-datasets-tenx_pbmc4k.rds")
```


### Feature selection


Read files from clusters
```{r}
sce <- readRDS("data/05.integrating-datasets-sce.rds")
tenx_pbmc3k <- readRDS("data/05.integrating-datasets-tenx_pbmc3k.rds")
tenx_pbmc4k <- readRDS("data/05.integrating-datasets-tenx_pbmc4k.rds")
```

Identifying the subset of the clean matrix to work with.

We order genes by decreasing biological component. We will be 
using this information later when performing feature selection
prior to running mnnCorrect().

```{r}
fit_sce <- trendVar(sce, use.spikes=FALSE) 
dec_sce <- decomposeVar(sce, fit_sce)
dec_sce$Symbol_TENx <- rowData(sce)$Symbol_TENx
dec_sce <- dec_sce[order(dec_sce$bio, decreasing = TRUE), ]

plot(dec_sce$mean, dec_sce$total, 
     xlab="Mean log-expression", 
     ylab="Variance of log-expression", pch=16)
curve(fit_sce$trend(x), col="dodgerblue", add=TRUE)
```

```{r}
fit_tenx_pbmc3k <- trendVar(tenx_pbmc3k, use.spikes=FALSE) 
dec_tenx_pbmc3k <- decomposeVar(tenx_pbmc3k, fit_tenx_pbmc3k)
dec_tenx_pbmc3k$Symbol_TENx <- rowData(tenx_pbmc3k)$Symbol_TENx
dec_tenx_pbmc3k <- dec_tenx_pbmc3k[order(dec_tenx_pbmc3k$bio, decreasing = TRUE), ]

plot(dec_tenx_pbmc3k$mean, dec_tenx_pbmc3k$total, 
     xlab="Mean log-expression", 
     ylab="Variance of log-expression", pch=16)
curve(fit_tenx_pbmc3k$trend(x), col="dodgerblue", add=TRUE)
```

```{r}
fit_tenx_pbmc4k <- trendVar(tenx_pbmc4k, use.spikes=FALSE) 
dec_tenx_pbmc4k <- decomposeVar(tenx_pbmc4k, fit_tenx_pbmc4k)
dec_tenx_pbmc4k$Symbol_TENx <- rowData(tenx_pbmc4k)$Symbol_TENx
dec_tenx_pbmc4k <- dec_tenx_pbmc4k[order(dec_tenx_pbmc4k$bio, decreasing = TRUE), ]

plot(dec_tenx_pbmc4k$mean, dec_tenx_pbmc4k$total, 
     xlab="Mean log-expression", 
     ylab="Variance of log-expression", pch=16)
curve(fit_tenx_pbmc4k$trend(x), col="dodgerblue", add=TRUE)
```


## Integrating Datasets

PCA plot of log normalized data with no batch correction
```{r}
sce <- runPCA(sce)
p_none <- plotReducedDim(sce, use_dimred="PCA",
               colour_by="Sample") + 
        ggtitle("Without batch correction")
p_none

```

PCA plot after using `removeBatchEffect()` from the `limma` package. 
```{r}
corrected <- limma::removeBatchEffect(logcounts(sce), batch=sce$Sample)
assay(sce, "corrected_logcounts_limma") <- corrected
rm(corrected)

p_limma <- plotPCASCE(sce, by_exprs_values = "corrected_logcounts_limma", 
        colour_by="Sample") + 
        ggtitle("Batch correction with limma")
p_limma

```

PCA plot after using `fastMNN()` from the `scran`

```{r}
universe <- intersect(rownames(dec_tenx_pbmc3k), rownames(dec_tenx_pbmc4k))
mean.bio <- (dec_tenx_pbmc3k[universe,"bio"] + dec_tenx_pbmc4k[universe,"bio"])/2
chosen <- universe[mean.bio > 0]
length(chosen)
```

We also rescale each batch to adjust for differences in 
sequencing depth between batches. The `multiBatchNorm()` function
recomputes log-normalized expression values after adjusting the
size factors for systematic differences in coverage between 
`SingleCellExperiment` objects. (Keep in mind that the previously 
computed size factors only remove biases between cells within 
a single batch.) This improves the quality of the correction by
removing one aspect of the technical differences between batches.

```{r}
rescaled <- multiBatchNorm(tenx_pbmc3k[universe,], tenx_pbmc4k[universe,])
rescaled_tenx_pbmc3k <- rescaled[[1]]
rescaled_tenx_pbmc4k <- rescaled[[2]]
```


```{r}

out <- fastMNN(rescaled_tenx_pbmc3k[chosen, ], 
               rescaled_tenx_pbmc4k[chosen, ], 
               k=20, d=50, approximate=TRUE)
dim(out$corrected)

sce_mnn <- SingleCellExperiment(
        assays=list(counts=cbind(counts(rescaled_tenx_pbmc3k[chosen, ]),
                                 counts(rescaled_tenx_pbmc4k[chosen, ])), 
                    logcounts=cbind(logcounts(rescaled_tenx_pbmc3k[chosen, ]), 
                                    logcounts(rescaled_tenx_pbmc4k[chosen, ]))), 
                     rowData = rowData(rescaled_tenx_pbmc3k[chosen, ]),
                     colData = rbind(colData(rescaled_tenx_pbmc3k[chosen, ]), 
                                     colData(rescaled_tenx_pbmc4k[chosen, ])))
reducedDim(sce_mnn, "corrected_logcounts_fastMNN") <- out$correct

p_mnn <- plotPCASCE(sce_mnn, 
                    by_exprs_values = "corrected_logcounts_fastMNN", 
                    colour_by="Sample") + 
        ggtitle("With batch correction using MNN")
```

Plot no batch correction to correcting batch effects with MNN
```{r}
library(cowplot)
pdf("data_analysis/figs/05.integrating-figure.pdf", width=9, height=4)
p <- plot_grid(p_none + theme(legend.position = "none"),
               p_mnn + theme(legend.position = "none"), ncol=2, labels = LETTERS[1:2])
legend <- get_legend(p_none + theme(legend.position = "bottom"))
p2 <- plot_grid(p,legend, rel_heights = c(2,.15), ncol=1) # c(1.25, .23))
title <- ggdraw() + draw_label("Correcting batch effects in single-cell RNA-sequencing data using MNN", fontface='bold')
plot_grid(title, p2, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
dev.off()

```



## Session Info

```{r}

```
