# Large Scale Single Cell Analyses

_Author_: Davide Risso

## Introduction

The purpose of this case study is to demonstrate how to use
R/Bioconductor packages to idenify clusters of cells from
scRNA-seq data with a large number cells. 

## Learning objectives

* In bullet form, illustrate what the main outputs are of this vignette
* A couple of bullet points are fine

## Package requirements

Packages that are required to go through the vignette and how to install them, should largely be through BiocManager interface.

```{r preamble, eval=FALSE}
BiocManager::install(c('scater', 'scran'))
```

```{r}
library(scater)
library(scran)
```

## Data

### Description of Data

Here the data should be loaded if in package form and briefly explored to illustrate the contents of the data. 

### Loading the Data

If data comes from "raw" form, then a simple pipeline for creating the processed data should be illustrated as succinctly as possible.

## Preprocessing

All steps required to preprocess the data into the clean expression matrix, split it up into subsections as necessary. Subsection splits should be informed by the paper (sub)sections. This is all steps prior to the main strategy of interest.

### Cell and gene quality control

Removal of "bad cells", low abundance genes. 

### Normalization

Application of normalization approach, even if its just a log normalization on the count data. Cell cycle normalization may be applicable here.

### Feature selection

Identifying the subset of the clean matrix to work with.

## Session Info

```{r}

```
